---
title: "bGMYC4 Interactive Workflow / Интерактивный рабочий процесс bGMYC4"
author: "Dmitry Karabanov & Qwen3.6-Plus"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bGMYC4 Interactive Workflow / Интерактивный рабочий процесс bGMYC4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# EN: Disable execution for CRAN checks to prevent timeouts on interactive MCMC.
# RU: Отключено выполнение для проверки CRAN, чтобы избежать таймаутов.
# Для локального запуска: измените eval = FALSE на eval = TRUE или запускайте блоки вручную.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 5,
  warning = FALSE, message = FALSE,
  eval = FALSE
)
```

## 🌍 Introduction / Введение    
**EN**: This vignette documents the complete interactive workflow for Bayesian species delimitation using bGMYC4. It guides you through data selection (built-in east10 or custom BEAST2 files), tree preprocessing, interactive MCMC diagnostics with parameter tuning, multi-tree uncertainty pooling, and final species hypothesis extraction. The workflow is designed for RStudio console execution.    
**RU**: Эта виньетка описывает полный интерактивный рабочий процесс байесовской делимитации видов с помощью bGMYC4. Она проведёт вас через выбор данных (встроенный east10 или собственные файлы BEAST2), предобработку деревьев, интерактивную диагностику MCMC с настройкой параметров, усреднение филогенетической неопределённости и извлечение гипотез о видах. Рабочий процесс оптимизирован для выполнения в консоли RStudio.    
---

## 1. Setup & Helper Functions / Настройка окружения и вспомогательные функции    
**EN:** Load required libraries and define safe input handlers. The workflow automatically cleans BEAST2 annotations, checks ultrametricity, and validates parameter bounds.    
**RU:** Загрузка библиотек и определение безопасных обработчиков ввода. Рабочий процесс автоматически очищает аннотации BEAST2, проверяет ультраметричность и валидирует границы параметров.    

```{r setup}
compiler::enableJIT(3) # JIT acceleration recommendation
library(ape)
library(bGMYC4)

# EN: Safe numeric input with bounds checking / Безопасный ввод чисел с проверкой границ    
safe_num <- function(prompt, default, min_v = NULL, max_v = NULL, hint = "") {
  if (!interactive()) return(default)
  inp <- readline(sprintf("%s [%s] %s: ", prompt, default, hint))
  if (inp == "") return(default)
  val <- suppressWarnings(as.numeric(inp))
  if (is.na(val)) {
    cat("  ⚠️ Invalid/Неверный ввод. Оставляю текущее.\n")
    return(default)
  }
  if (!is.null(min_v) && val < min_v) {
    cat(sprintf("  ⚠️ < min (%s). Using %s.\n", min_v, min_v))
    return(min_v)
  }
  if (!is.null(max_v) && val > max_v) {
    cat(sprintf("  ⚠️ > max (%s). Using %s.\n", max_v, max_v))
    return(max_v)
  }
  return(val)
}

# EN: Read Nexus files while stripping BEAST2 metadata / Чтение Nexus с удалением метаданных BEAST2    
read_beast_nexus <- function(file_path) {
  cat(sprintf("📥 Reading / Читаю: %s\n", basename(file_path)))
  lines <- readLines(file_path, warn = FALSE)
  lines <- gsub("\\[&[^]]*\\]", "", lines)  # Strip [&...] tags
  tmp <- tempfile(fileext = ".trees")
  writeLines(lines, tmp)
  tr <- read.nexus(tmp)
  unlink(tmp)
  return(tr)
}

# EN: Enforce ultrametricity with float rounding / Обеспечение ультраметричности    
fix_ultrametric <- function(tr) {
  if (!is.ultrametric(tr)) {
    cat("⚠️ Not ultrametric / Не ультраметрично. Округляю длины ветвей / Rounding edges...\n")
    tr$edge.length <- round(tr$edge.length, 8)
  }
  if (!is.ultrametric(tr)) stop("❌ Still not ultrametric / Всё ещё не ультраметрично. Проверьте экспорт BEAST2.")
  return(tr)
}
```

---

## 2. Data Selection & Tree Preprocessing / Выбор данных и предобработка деревьев    
**EN:** Choose between the simulated `east10` dataset or your own BEAST2 files. The script guides you through rooting, outgroup removal, and ultrametricity verification.  
**RU:** Выбор между симулированным датасетом `east10` или вашими файлами BEAST2. Скрипт проведёт через укоренение, удаление аутгрупп и проверку ультраметричности.    

```{r preprocessing}
cat("Select data source / Выберите источник данных:\n")
cat("  1. Built-in 'east10' / Встроенный датасет 'east10'\n")
cat("  2. Custom BEAST2 files / Собственные файлы\n")
choice <- safe_num("Enter 1 or 2 / Введите 1 или 2", 1, min_v = 1, max_v = 2)

if (choice == 1) {
  data(east10)
  tree_consensus <- east10[[1]]
  all_trees <- east10
} else {
  if (interactive()) {
    cat("\n📂 Select directory / Выберите директорию:\n")
    dir_path <- utils::choose.dir(default = getwd(), caption = "Select trees folder")
    if (dir_path == "") stop("❌ Directory not selected / Директория не выбрана.")
    setwd(dir_path)
    tree_consensus <- read_beast_nexus(utils::choose.files(caption = "Select .tree file"))
    all_trees      <- read_beast_nexus(utils::choose.files(caption = "Select .trees file"))
  } else {
    stop("Для собственных файлов укажите пути к .tree и .trees вручную.")
  }
}

ntips <- length(tree_consensus$tip.label)
cat(sprintf("✅ Trees loaded / Загружено деревьев: %d | Tips / Таксонов: %d\n", length(all_trees), ntips))

# EN: Rooting check & outgroup removal / Проверка укоренения и удаление аутгрупп    
if (interactive()) {
  cat("\n🔍 Rooting check / Проверка укоренения:\n")
  if (!all(sapply(all_trees, is.rooted))) {
    og_str <- readline("Enter outgroup(s) / Введите аутгруппу(ы) (comma-separated): ")
    outgroups <- trimws(strsplit(og_str, ",")[[1]])
    all_trees      <- lapply(all_trees, function(t) if (!is.rooted(t)) root(t, outgroups[1], resolve.root = TRUE) else t)
    tree_consensus <- if (!is.rooted(tree_consensus)) root(tree_consensus, outgroups[1], resolve.root = TRUE) else tree_consensus
  }
  
  cat("\n🔪 Remove taxa from analysis? / Удалить таксоны из анализа? (y/n): ")
  if (tolower(readline("")) %in% c("y", "yes", "да", "")) {
    og_str <- readline("Enter taxa to drop / Введите таксоны для удаления (comma-separated): ")
    drop_list <- trimws(strsplit(og_str, ",")[[1]])
    tree_consensus <- drop.tip(tree_consensus, drop_list)
    all_trees      <- lapply(all_trees, drop.tip, tip = drop_list)
    class(all_trees) <- "multiPhylo"
    ntips <- length(tree_consensus$tip.label)
    cat(sprintf("✅ Tips remaining / Осталось таксонов: %d\n", ntips))
  }
}

# EN: Final ultrametric check & tree sampling / Финальная проверка и выборка    
tree_consensus <- fix_ultrametric(tree_consensus)
all_trees      <- lapply(all_trees, fix_ultrametric)
class(all_trees) <- "multiPhylo"

n_sample <- safe_num("How many trees to sample? / Сколько деревьев взять? (10-100 recommended)", 10, min_v = 1, max_v = length(all_trees))
trees_sample <- sample(all_trees, n_sample)
class(trees_sample) <- "multiPhylo"
cat(sprintf("✅ Sampled / Выбрано: %d trees\n", n_sample))
```

---

## 3. Interactive Diagnostics & Parameter Tuning / Интерактивная диагностика и настройка
**EN:** The core diagnostic loop. It runs bgmyc.singlephy, evaluates acceptance rates and logposterior stationarity, plots trace graphs, and prompts you to adjust MCMC parameters before scaling to multiple trees.    
**RU:** Основной диагностический цикл. Запускает bgmyc.singlephy, оценивает acceptance rates и стационарность логарифма правдоподобия, строит графики и предлагает настроить параметры MCMC перед многодеревным анализом.

```{r diagnostics-loop}
# EN: Initial parameters / Начальные параметры    
prm <- list(
  mcmc = 5000, burnin = 1000, thinning = 10,
  py1 = 0, py2 = 1.5, pc1 = 0, pc2 = 2.0,
  t1 = 2, t2 = min(35, ntips - 5),
  scale = c(25, 10, 5), start = c(1, 0.5, floor(ntips/3))
)
prm$start[3] <- max(prm$t1 + 1, min(prm$start[3], prm$t2 - 1)) # Safety clamp

# EN: Diagnostic printer & plotter / Функция диагностики и графиков
show_diag <- function(res, prm) {
  cat("\n📊 FULL DIAGNOSTICS / ПОЛНАЯ ДИАГНОСТИКА\n")
  cat(sprintf("⚙️  Parameters: mcmc=%d | burnin=%d | thin=%d\n", prm$mcmc, prm$burnin, prm$thinning))
  acc <- res$accept
  cat(sprintf("📈 Acceptance rates: py=%.2f | pc=%.2f | t=%.2f [Optimal/Оптимум: 0.20–0.40]\n", acc[1], acc[2], acc[3]))

  # 🔢 ESS Calculation (Effective Sample Size)
  if (requireNamespace("mcmcse", quietly = TRUE)) {
    ess_vals <- sapply(1:4, function(col) round(mcmcse::ess(res$par[, col])))
    cat(sprintf("🔢 ESS: py=%d | pc=%d | th=%d | logL=%d [Optimal/Оптимум: >200]\n", 
                ess_vals[1], ess_vals[2], ess_vals[3], ess_vals[4]))
  } else {
    cat("⚠️  ESS: package 'mcmcse' not installed. Run install.packages('mcmcse') for full diagnostics.\n")
    ess_vals <- NULL
  }

  par_vec <- res$par[, 4]
  q <- split(par_vec, cut(seq_along(par_vec), 4, labels = FALSE))
  cv <- sd(sapply(q, var)) / mean(sapply(q, var))
  cat(sprintf("📉 Stationarity (CV<0.6): %s\n", ifelse(cv < 0.6, "✅ YES/ДА", "⚠️ CHECK/ПРОВЕРЬТЕ")))

  cat("\n💡 Recommendations / Рекомендации:\n")
  if (acc[1] > 0.50) cat("   ↑ scale[1] or ↑ py2 (py too conservative)\n")
  if (acc[1] < 0.15) cat("   ↓ scale[1] (py too bold)\n")
  if (acc[2] > 0.50) cat("   ↑ scale[2]\n")
  if (acc[2] < 0.15) cat("   ↓ scale[2]\n")
  if (!is.null(ess_vals) && any(ess_vals < 200)) {
    cat("   ↑ mcmc or ↓ thinning (ESS < 200: цепь требует больше независимых шагов)\n")
  }
  if (cv >= 0.6) cat("   ↑ burnin to 30% mcmc or ↑ mcmc (chain not stationary)\n")
  if (cv < 0.6 && all(acc > 0.2 & acc < 0.45)) cat("   ✅ Parameters optimal. Ready for multi-tree.\n")

  cat("\n🖼️  Opening trace plots... / Открываю графики сходимости...\n")
  plot(res)
  cat("👀 Close plot window when done. / Закройте окно графиков, когда закончите.\n")
}

# EN: Interactive tuning loop / Интерактивный цикл настройки    
cat("\n🔄 Starting diagnostic loop / Запускаю диагностику...\n")
repeat {
  cat("🚀 Running single-tree MCMC / Запуск MCMC на одном дереве...\n")
  res <- bgmyc.singlephy(
    phylo = tree_consensus, mcmc = prm$mcmc, burnin = prm$burnin, thinning = prm$thinning,
    py1 = prm$py1, py2 = prm$py2, pc1 = prm$pc1, pc2 = prm$pc2, t1 = prm$t1, t2 = prm$t2,
    scale = prm$scale, start = prm$start
  )
  show_diag(res, prm)
  
  cat("❓ Use these parameters for multi-tree analysis? / Использовать эти параметры? (y/n): ")
  if (!interactive()) break
  ans <- tolower(trimws(readline("")))
  if (ans %in% c("y", "да", "")) break
  
  cat("\n🔧 Adjust parameters (Enter=keep current) / Настройка параметров (Enter=оставить текущее)\n")
  prm$mcmc     <- safe_num("mcmc", prm$mcmc, min_v = 1000, hint = "[↑=more accurate]")
  prm$burnin   <- safe_num("burnin", prm$burnin, min_v = 100, max_v = prm$mcmc - 100, hint = "[20-30% of mcmc]")
  prm$thinning <- safe_num("thinning", prm$thinning, min_v = 1, hint = "[↓ autocorrelation]")
  prm$py2      <- safe_num("py2", prm$py2, min_v = prm$py1 + 0.1, hint = "[≤1.5 recommended]")
  prm$pc2      <- safe_num("pc2", prm$pc2, min_v = prm$pc1 + 0.1, hint = "[≤2.0]")
  prm$t2       <- safe_num("t2", prm$t2, max_v = ntips - 2, hint = "[< number of tips]")
  prm$scale    <- c(safe_num("scale[1]", prm$scale[1], min_v = 1),
                    safe_num("scale[2]", prm$scale[2], min_v = 1),
                    safe_num("scale[3]", prm$scale[3], min_v = 1))
  prm$start    <- c(safe_num("start[1]", prm$start[1], min_v = 0),
                    safe_num("start[2]", prm$start[2], min_v = 0),
                    safe_num("start[3]", prm$start[3], min_v = prm$t1 + 1, max_v = prm$t2 - 1))
  prm$start[3] <- max(prm$t1 + 1, min(prm$start[3], prm$t2 - 1))
}
```

---

## 4. Multi-Tree Analysis & Final Output / Многодеревный анализ и финальный вывод    
**EN:** Once diagnostics are stable, bgmyc.multiphylo runs on the sampled trees. The workflow then builds a conspecificity probability matrix, visualizes it as a heatmap, and extracts species clusters at your chosen threshold.    
**RU:** После стабилизации диагностики bgmyc.multiphylo запускается на выбранных деревьях. Строится матрица вероятностей конспецифичности, визуализируется как тепловая карта, и извлекаются кластеры видов при выбранном пороге.    

```{r final-analysis}
cat("\n🌲 Running multi-tree analysis / Запуск анализа на множестве деревьев...\n")
result_multi <- bgmyc.multiphylo(
  multiphylo = trees_sample, mcmc = prm$mcmc, burnin = prm$burnin, thinning = prm$thinning,
  py1 = prm$py1, py2 = prm$py2, pc1 = prm$pc1, pc2 = prm$pc2,
  t1 = prm$t1, t2 = prm$t2, scale = prm$scale, start = prm$start
)

# 📐 Gelman-Rubin (R̂) convergence check across tree chains
if (requireNamespace("mcmcse", quietly = TRUE) && length(result_multi) > 1) {
  cat("\n📐 Gelman-Rubin R̂ convergence check / Проверка сходимости R̂...\n")
  chains_list <- lapply(result_multi, function(res) res$par[, 3]) # threshold parameter
  names(chains_list) <- paste0("Tree_", seq_along(result_multi))
  
  gr_result <- tryCatch(mcmcse::gelman(chains_list), error = function(e) NULL)
  if (!is.null(gr_result)) {
    rhat <- round(gr_result$Rhat, 3)
    cat(sprintf("🔍 Gelman-Rubin R̂ (threshold): %.3f [Optimal/Оптимум: < 1.05]\n", rhat))
    if (rhat > 1.05) {
      cat("⚠️  R̂ > 1.05: Chains show divergence. Increase mcmc/burnin or check tree topologies.\n")
    } else {
      cat("✅ Chains converged stably across posterior trees / Цепи сошлись стабильно.\n")
    }
  }
}

cat("\n🌡️  Building heatmap / Строю тепловую карту...\n")
prob_mat <- spec.probmat(result_multi)
plot(prob_mat, tree_consensus)  # Ordered by consensus topology / Упорядочено по консенсусу

cat("\n🔍 Delimitation (PP > 0.05) / Делимитация (PP > 0.05)...\n")
out <- bgmyc.point(prob_mat, ppcutoff = 0.05)
cat("✅ Clusters / Кластеров:", length(out), "\n")
for (i in seq_along(out)) {
  cat(paste0("  ", i, ": ", paste(out[[i]][1:4], collapse = ", "),
             if (length(out[[i]]) > 4) "...\n" else "\n"))
}

# EN: Export cluster probabilities to CSV / Экспорт вероятностей кластеров в CSV    
spec_out <- bgmyc.spec(result_multi)
write.csv(spec_out$specprobs, file = "bGMYC_clusters.csv", row.names = FALSE, fileEncoding = "UTF-8")

save(prm, res, result_multi, prob_mat, out, file = "bGMYC_final_results.RData")
cat("💾 Results saved / Результаты сохранены.\n🎉 Analysis complete / Анализ завершён!\n")
```

---

## ⚡ Performance & Export / Производительность и экспорт    
**EN:** bgmyc.multiphylo() automatically parallelizes across physical CPU cores. To limit cores (e.g., to 6), run before analysis: options(mc.cores = 6).    
**RU:** bgmyc.multiphylo() автоматически использует все физические ядра. Чтобы ограничить число ядер (например, до 6), выполните перед запуском: options(mc.cores = 6).    

```{r performance-export}
# EN: Multi-tree analysis now runs in parallel automatically across all CPU cores / Многодеревный анализ теперь автоматически запускается параллельно на всех ядрах CPU.    
# EN: Export clusters to CSV (Excel-compatible) / Экспорт кластеров в CSV (совместимо с Excel)     
# EN: Optional: Validate model fit by comparing Yule vs Coalescent rates / Опционально: проверь соответствие модели, сравнив скорости видообразования и коалесценции    

---

## 📊 Parameter Reference Guide / Справочник по параметрам
| Parameter / Параметр | Role / Роль | Recommended Range | Biological/Statistical Notes / Примечания |
|:---|:---|:---|:---|
| `mcmc` | Chain length / Длина цепи | `10k` (test), `50k+` (final) | Longer chains improve posterior resolution. / Длинные цепи улучшают апостериорную оценку. |
| `burnin` | Warm-up / Разогрев | `20–30%` of `mcmc` | Discards non-stationary start. Increase if logposterior drifts. / Отбрасывает неустановившуюся фазу. |
| `thinning` | Sampling interval / Интервал выборки | `10–50` | Reduces autocorrelation & RAM. Higher for long chains. / Снижает автокорреляцию и нагрузку на память. |
| `py1`, `py2` | Yule rate prior / Априор видообразования | `0`, `0.5–1.5` | Model: `λ ∝ n^py`. `py > 1.5` blurs Yule/Coalescent boundary. / >1.5 размывает границу модели. |
| `pc1`, `pc2` | Coalescent prior / Априор коалесценции | `0`, `1.0–2.0` | Models `Ne` change. `pc < 1` → decline, `pc > 1` → growth. / Моделирует динамику эффективного размера популяции. |
| `t1`, `t2` | Threshold prior (species count) / Априор числа видов | `2`, `min(35, ntips-5)` | Must be `< ntips`. Auto-capped to prevent crashes. / Строго `< числа таксонов`. |
| `scale` | Proposal step widths / Ширина предложений MCMC | `c(20–30, 10–15, 3–7)` | Higher = more conservative. Tune via acceptance rates. / Выше = консервативнее. Настраивается по acceptance rates. |
| `ppcutoff` | Lumping threshold / Порог объединения | `0.05` (variable), `0.95` (strict) | Low = captures high intraspecific variation/ILS. / Низкий = учитывает высокую внутривидовую изменчивость. |

---

## 🔍 Convergence Diagnostics Guide / Руководство по диагностике сходимости    
**EN:** After the diagnostic run, evaluate two metrics:    
1. **Acceptance Rates**: `0.20–0.40` is optimal. `<0.15` → decrease `scale`. `>0.50` → increase `scale`.    
2. **Trace Plots ("Fuzzy Caterpillars")**: After burnin, parameters should oscillate horizontally around a stable mean. Upward/downward trends → increase burnin or mcmc. Sticky steps → scale too high. Smooth/lazy curves → scale too low. Hitting bounds → widen priors.    

**RU:** После диагностического запуска проверьте два показателя:    
1. **Acceptance Rates**: Оптимум `0.20–0.40`. `<0.15` → уменьшите `scale`. `>0.50` → увеличьте `scale`.    
2. **Графики ("пушистые гусеницы")**: После burnin параметры должны колебаться горизонтально вокруг стабильного среднего. Тренды вверх/вниз → увеличьте `burnin` или `mcmc`. Ступеньки → `scale` слишком высок. Гладкие кривые → `scale` слишком низок. Прилипание к границам → расширьте априоры.    

---

## 📎 Notes for CRAN & Local Use / Примечания для CRAN и локального запуска    
**EN:**
- `eval = FALSE` in code chunks prevents CRAN from timing out during automated checks.    
- To run interactively: change `eval = FALSE` to `eval = TRUE` in the first chunk, or simply copy-paste chunks into RStudio console and execute sequentially.    
- Always run diagnostics on 1 tree before scaling to `multiPhylo`.    

**RU:**
- `eval = FALSE` предотвращает таймауты при автоматической проверке CRAN.    
- Для локального запуска: измените `eval = FALSE` на `eval = TRUE` в первом блоке, или копируйте блоки в консоль RStudio и запускайте последовательно.    
- Всегда запускайте диагностику на 1 дереве перед переходом к `multiPhylo`.    

---
**References / Ссылки:** Pons et al. 2006 *Syst. Biol.* 55:595; Reid & Carstens 2012 *Mol. Ecol. Res.* 12:446.

`vignette("bGMYC4-interactive", package = "bGMYC4")`
